\texttt{davos}: a Python package ``smuggler'' for constructing lightweight reproducible notebooks

### 2022-10-28 paper/main.tex
updates through section 2

\subsection{Software functionalities}
% Present the major functionalities of the software.
\subsubsection{The \texttt{smuggle} statement}\label{subsec:smuggle}
Importing \texttt{davos} in an IPython notebook appears to enable an additional Python keyword: ``\texttt{smuggle}'' (see Section~\ref{subsec:implementation} for details on how this works).
The \texttt{smuggle} statement can be used as a drop-in replacement for Python's built-in \texttt{import} statement to load libraries, modules, and other objects into the current namespace.
However, whereas \texttt{import} will fail if the requested package is not installed locally, \texttt{smuggle} statements can handle missing packages on the fly.
If a smuggled package does not exist in the local environment, \texttt{davos} will install it automatically, expose its contents to Python's \texttt{import} machinery, and load it into the namespace for immediate use.

\subsubsection{The onion comment}\label{subsec:onion}
For greater control over the behavior of \texttt{smuggle} statements, \texttt{davos} defines an additional construct called the ``onion comment.'' An onion comment is a special type of inline comment that may be placed on a line containing a \texttt{smuggle} statement to customize how \texttt{davos} searches for the smuggled package locally and, if necessary, downloads and installs it. Onion comments follow a simple format based on the ``type comment'' syntax introduced in PEP 484~\cite{vanREtal14}, and are designed to make managing packages with \texttt{davos} intuitive and familiar. To construct an onion comment, users provide the name of the installer program (e.g., \texttt{pip}) and the same arguments one would use to manually install the package as desired via the command line:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet1}
\end{center}
Occasionally, a package's distribution name (i.e., the name used when installing it) may differ from its top-level module name (i.e., the name used when importing it). In such cases, an onion comment can be used to ensure \texttt{davos} installs the proper distribution if the smuggled package can't be found locally:
%Onion comments are useful in such cases for ensuring \texttt{davos} installs the proper distribution if the smuggled package isn't installed locally:
%Onion comments are useful when smuggling a package whose distribution name (i.e., the name used when installing it) differs from its top-level module name (i.e., the name used when importing it). Onion comments are useful in such cases for 
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet2}
\end{center}
However, the most powerful use of the onion comment is making \texttt{smuggle} statements \textit{version-sensitive}. If an onion comment includes a version specifier~\cite{CoghStuf13}, \texttt{davos} will ensure that the version of the package loaded into the notebook matches the specific version requested, or satisfies the given version constraints. If the smuggled package exists locally, \texttt{davos} will extract its version info from its metadata and compare it to the specifier provided. If the two are incompatible (or no local installation is found), \texttt{davos} will install and load a suitable version of the package instead:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet3}
\end{center}
Onion comments can also be used to smuggle specific VCS references (e.g., Git~\cite{TorvHama05}  branches, commits, tags, etc.):
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet4}
\end{center}
\texttt{davos} processes onion comments internally before forwarding arguments to the installer program. In addition to preventing onion comments from being used as a vehicle for shell injection attacks, this allows \texttt{davos} to take certain logical actions when particular arguments are passed. For example, the \texttt{-I}/\newline\texttt{--ignore-installed}, \texttt{-U}/\texttt{--upgrade}, and \texttt{--force-reinstall} flags will all cause \texttt{davos} to skip searching for a smuggled package locally before installing a new copy:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet5}
\end{center}
Similarly, passing \texttt{--no-input} will temporarily enable \texttt{davos}'s non-interactive mode (see Section~\ref{subsec:onion}), and installing a smuggled package into \texttt{<dir>} with \texttt{--target <dir>} will cause \texttt{dir} to be prepended to the module search path (\texttt{sys.path}), if necessary, so the package can be imported

@@

\subsection{Implementation details}\label{subsec:implementation}
Functionally, importing \texttt{davos} appears to define ``\texttt{smuggle}'' as a Python keyword, similar to ``\texttt{import}'', ``\texttt{def}'', or ``\texttt{return}''. It also appears to cause comments to be parsed, and their contents potentially able to affect code behavior, which they normally are not. However, \texttt{davos} doesn't actually modify the rules of Python's parser or lexical analyzer---in fact, modifying the Python grammar isn't possible at runtime, as doing so would require rebuilding the interpreter. Instead, \texttt{davos} leverages the IPython notebook backend to implement the \texttt{smuggle} statement and onion comment via a combination of namespace injections and its own (far simpler) custom parser.

The \texttt{smuggle} keyword can be enabled and disabled at any time by ``activating'' and ``deactivating'' \texttt{davos} (see Section~\ref{subsec:config}, above). When \texttt{davos} is first imported, it is activated automatically. Activating \texttt{davos} triggers two actions: (1) the \texttt{smuggle()} function is injected into the IPython user namespace, and (2) the \texttt{davos} parser is registered as a custom IPython input transformer. IPython preprocesses all executed code as plain text before it is sent to the Python parser, in order to handle special constructs like \texttt{\%magic} and \texttt{!shell} commands. \texttt{davos} hooks into this process to transform \texttt{smuggle} statements into syntactically valid Python code. The \texttt{davos} parser uses a complex regular expression~\cite{Thom68} to match lines of code containing \texttt{smuggle} statements (and, optionally, onion comments), extract relevant information from their text, and replace them with equivalent calls to the \texttt{smuggle()} function. For example, if a user runs a notebook cell containing
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet6}
\end{center}
the code that is actually executed by the Python interpreter would be
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet7}
\end{center}
Because the \texttt{smuggle()} function is defined in the notebook namespace, it is also possible (though never necessary) to call it directly. Deactivating \texttt{davos} will delete the name ``\texttt{smuggle}'' from the namespace, unless its value has been overwritten and no longer refers to the \texttt{smuggle()} function. It will also deregister the \texttt{davos} parser from the set of input transformers run when each notebook cell is executed. While the overhead added by the \texttt{davos} parser is de minimis, this may be useful, for example, when optimizing or precisely profiling code.

>>>>

\subsection{Software functionalities}
% Present the major functionalities of the software.
\subsubsection{The \texttt{smuggle} statement}\label{subsec:smuggle}
Importing \texttt{davos} in an IPython notebook enables an additional Python keyword: ``\texttt{smuggle}'' (see Section~\ref{subsec:implementation} for details on how this works).
The \texttt{smuggle} statement can be used as a drop-in replacement for Python's built-in \texttt{import} statement to load libraries, modules, and other objects into the current namespace.
However, whereas \texttt{import} will fail if the requested package is not installed locally, \texttt{smuggle} statements can handle missing packages on the fly.
If a smuggled package does not exist in the local environment, \texttt{davos} will install it automatically, expose its contents to Python's \texttt{import} machinery, and load it into the namespace for immediate use.

\subsubsection{The onion comment}\label{subsec:onion}
For greater control over the behavior of \texttt{smuggle} statements, \texttt{davos} defines an additional construct called the ``onion comment.'' An onion comment is a special type of inline comment that may be placed on a line containing a \texttt{smuggle} statement to customize how \texttt{davos} searches for the smuggled package locally and, if necessary, downloads and installs it. Onion comments follow a simple format based on the ``type comment'' syntax introduced in PEP 484~\cite{vanREtal14}, and are designed to make managing packages with \texttt{davos} intuitive and familiar. To construct an onion comment, users provide the name of the installer program (e.g., \texttt{pip}) and the same arguments one would use to manually install the package as desired via the command line:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet1}
\end{center}
Occasionally, a package's distribution name (i.e., the name used when installing it) may differ from its top-level module name (i.e., the name used when importing it). In such cases, an onion comment can be used to ensure \texttt{davos} installs the proper distribution if the smuggled package can't be found locally:
%Onion comments are useful in such cases for ensuring \texttt{davos} installs the proper distribution if the smuggled package isn't installed locally:
%Onion comments are useful when smuggling a package whose distribution name (i.e., the name used when installing it) differs from its top-level module name (i.e., the name used when importing it). Onion comments are useful in such cases for
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet2}
\end{center}
However, the most powerful use of the onion comment is making \texttt{smuggle} statements \textit{version-sensitive}. If an onion comment includes a version specifier~\cite{CoghStuf13}, \texttt{davos} will ensure that the version of the package loaded into the notebook matches the specific version requested, or satisfies the given version constraints. If the smuggled package exists locally, \texttt{davos} will extract its version info from its metadata and compare it to the specifier provided. If the two are incompatible (or no local installation is found), \texttt{davos} will install and load a suitable version of the package instead:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet3}
\end{center}
Onion comments can also be used to smuggle specific VCS references (e.g., Git~\cite{TorvHama05}  branches, commits, tags, etc.):
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet4}
\end{center}
\texttt{davos} processes onion comments internally before forwarding arguments to the installer program. In addition to preventing onion comments from being used as a vehicle for shell injection attacks, this allows \texttt{davos} to take certain logical actions when particular arguments are passed. For example, the \texttt{-I}/\newline\texttt{--ignore-installed}, \texttt{-U}/\texttt{--upgrade}, and \texttt{--force-reinstall} flags will all cause \texttt{davos} to skip searching for a smuggled package locally before installing a new copy:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet5}
\end{center}
Similarly, passing \texttt{--no-input} will temporarily enable \texttt{davos}'s non-interactive mode (see Section~\ref{subsec:config}), and installing a smuggled package into \texttt{<dir>} with \texttt{--target <dir>} will cause \texttt{dir} to be prepended to the module search path (\texttt{sys.path}), if necessary, so the package can be imported

@@

\subsection{Implementation details}\label{subsec:implementation}
Functionally, importing \texttt{davos} appears to make ``\texttt{smuggle}'' a valid Python keyword, similar to standard keywords like ``\texttt{import}'', ``\texttt{def}'', or ``\texttt{return}''. It also appears to fundamentally change how Python treats comments, suddenly allowing them to potentially influence code behavior at runtime if they conform to a particular syntax. However, \texttt{davos} doesn't actually modify the rules of Python's parser or lexical analyzer in order to accomplish this---in fact, modifying the Python grammar isn't possible at runtime, as doing so would require rebuilding the interpreter. Instead, \texttt{davos} leverages the IPython notebook backend to implement the \texttt{smuggle} statement and onion comment via a combination of namespace injections and its own (far simpler) custom parser.

The \texttt{smuggle} keyword can be enabled and disabled at any time by ``activating'' and ``deactivating'' \texttt{davos} (see Section~\ref{subsec:config}, above). When \texttt{davos} is first imported, it is activated automatically. Activating \texttt{davos} triggers two actions: (1) the \texttt{smuggle()} function is injected into the IPython user namespace, and (2) the \texttt{davos} parser is registered as a custom IPython input transformer. IPython preprocesses all executed code as plain text before it is sent to the Python compiler, in order to handle special constructs like \texttt{\%magic} and \texttt{!shell} commands. \texttt{davos} hooks into this process to transform \texttt{smuggle} statements into syntactically valid Python code. The \texttt{davos} parser uses a complex regular expression~\cite{Thom68} to match lines of code containing \texttt{smuggle} statements (and, optionally, onion comments), extract relevant information from their text, and replace them with equivalent calls to the \texttt{smuggle()} function. For example, if a user runs a notebook cell containing
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet6}
\end{center}
the code that is actually executed by the Python interpreter would be
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet7}
\end{center}
The call to the \texttt{smuggle()} function then carries out \texttt{davos}'s central logic of determining whether the smuggled package should be installed, doing so if necessary, and subsequently loading it into the namespace. This process is outlined in Figure \ref{fig:flow-chart}. Because the \texttt{smuggle()} function is defined in the notebook namespace, it is also possible (though never necessary) to call it directly. Deactivating \texttt{davos} will delete the name ``\texttt{smuggle}'' from the namespace, unless its value has been overwritten and no longer refers to the \texttt{smuggle()} function. It will also deregister the \texttt{davos} parser from the set of input transformers run when each notebook cell is executed. While the overhead added by the \texttt{davos} parser is minimal, this may be useful, for example, when optimizing or precisely profiling code.
